---
title: "Analysis of Limb Volume"
author: "Caleb Pollock"
date: '2023-01-11'
output: html_document
---
The data and code can be found at https://github.com/caleb-pollock/BIOL432_Assignment1

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```



To begin, we will import our data generator script which will create a data frame sampling 100 species names, limb widths, limb heights, and observers. This will then be output into a csv file known as measurements.csv

```{r}
source('dataGenerato.R')
```

Next, we use the csv file generated in the original script to add a new column which calculates the volume from the width and height values generated in the previous script. This includes using dplyrs pipe and mutate functions to add a new column, and we use the volume of a cylinder to calculate the volume of an organisms leg.This will then replace the measurements.csv file to include the new column

```{r}
source('volumeEstimato.R')
```

After executing both of the R scripts, we move onto loading the revised measurements.csv file.

```{r}
#creating a new dataframe containing the measurements.csv data and loading the revised measurements.csv
measurementDat <- read.csv("measurements.csv")
head(measurementDat)
```

We will now use dplyr to sort the data by species, than by observer, than by limb volume

```{r}
#apply the arrange function using the measurementDat dataframe and pipe to create a new dataframe which is arranged properly with Organism > Species > Volume
measurementDat <- measurementDat %>%
  arrange(Organism, Observer, Volume)
#you will now be able to see the data grouped by species, then by observer, and finally lowest to highest volume
head(measurementDat)
```
 
Following sorting the dataframe, we will now create a table showing the average volume for each species

```{r}
#using dplyr pipes, group_by() and sumarise() will allow use to create a table with average volume for each species. This will create a new dataframe called measurementTab that will be a table of the average volume for each species
measurementTab <- measurementDat %>%
  group_by(Organism) %>%
  summarize(AverageVol = mean(Volume))
head(measurementTab)
```
Following the average volume per species table, we will create a new table which sorts how many times an observer made observations on a particular organism.

```{r}
#using similar methods to the table above, we apply measurementDat to a new dataframe called speciesObsTab, which will be a table containing Organism, observer, and observations. group_by() will select the organism and observer to be included in the new table, and summarize() will return the current group size (n()) under a new column called Observations. 
speciesObsTab <- measurementDat %>%
  group_by(Organism, Observer) %>%
  summarize(Observations = n())
head(speciesObsTab)
```

